---
title: "Exploring Service Statistics and Effect of Fatigue in WTA Data"
# subtitle: "&#x2694;&#xFE0F; xaringan +<br/>&#x1F60E; xaringanthemer"  
author: 
  - "Andrew Noecker, Mykalyster Homberg, Phong Hoang"
  # - "Mykalyster Homberg"
  # - "Phong Hoang"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: FALSE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#dc4c4c",
  secondary_color = "#157394",
  inverse_header_color = "#FFFFFF"
)
```

```{r libraries & data, include = FALSE, warning = FALSE}
library(tidyverse)
library(ggthemes)
library(protoclust)
library(ggdendro)
library(gt)

wta <- read_csv("../Data/wta.csv")
duplicate_match <- read_csv("../Data/duplicate_match_wta.csv")
indiv_player_match <- read_csv("../Data/indiv_player_match_wta.csv")
player_serve_stats <- read_csv("../Data/player_serve_stats.csv")
final_minutes <- read_csv("../Data/final_minutes.csv")

```


## The Women's Tennis Association (WTA) Dataset

- **10,701 observations of 49 variables**
  - Courtesy of [Jeff Sackmann's tennis repository](https://github.com/JeffSackmann/tennis_wta)
  - One observation = one match between **2018** and **2022**

- **Key Variables**
  - Tournament / Match Information: name, date, location, surface
  - Player Information: seed, ranking, height
  - Match statistics: aces, double faults, other service statistics

<!-- - **Key match variables**: tourney_name, tourney_date, surface -->

```{r, echo = FALSE}
wta%>%
  select(tourney_name,tourney_date, surface, winner_name, loser_name, winner_rank, w_ace, w_df, w_bpSaved, w_1stIn)%>%
  head(n = 5)%>%
  gt()
# DT::datatable(fillContainer = FALSE, options = list(pageLength = 4))
```

---

## Data Transformations

- Created three datasets using original `wta` dataset

--

  1. `indiv_player_match`: data where each row is one player's performance in a match

--

  2. `player_serve_stats`: aggregated data of each player's cumulative statistics across all matches

--

  3. `final_minutes`: contains each tournament finalist and the minutes played prior to the final
  
--

- New Variables:
  
  - Number of sets
  
  - Winning Percentage
  
  - Minutes played leading up to a tournament final
  
  - Difference in minutes played between the two finalists prior to the final 

---

## <center> Winning players face fewer break points and save break points at better rates </center>

```{r, echo = FALSE, fig.height = 4}
bp_stats <- indiv_player_match%>%
  mutate(win = ifelse(result == "W", 1, 0))%>%
  group_by(player_name)%>%
  summarize(matches = n(),
            wins = sum(win, na.rm = TRUE),
            sum_faced = sum(bpFaced, na.rm = TRUE),
            sum_saved = sum(bpSaved, na.rm = TRUE),
            mean_faced = mean(bpFaced, na.rm = TRUE),
            mean_saved = mean(bpSaved, na.rm = TRUE))%>%
  mutate(prop_saved = sum_saved / sum_faced,
         win_pct = wins / matches)

p1 <- bp_stats%>%
  filter(matches >= 50)%>%
  ggplot(aes(x = prop_saved, y = win_pct))+
  geom_point(aes(color = mean_faced),alpha = 0.75, size = 4)+
  scale_color_gradient(limits = c(5, 12), low = "darkorange", high = "darkblue")+
  theme_bw()+
  labs(x = "Proportion of Break Points Saved",
       y = "Winning Percentage",
       color = "Average Break\nPoints Faced",
       # title = "Winning players face fewer break points and save break points at better rates",
       caption = "Minimum 50 matches played")+
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption.position = "plot")

p1
```

---

### Failing first serve results in worse point winning percentage
- 3 possible outcomes of serving: first serve in, second serve in, double faults

--

```{r, echo = FALSE, fig.height = 4}
# player_serve_stats <- read.csv("file.csv")
player_serve_stats %>%
  pivot_longer(cols = c(13:15)) %>%
  ggplot(aes(y = value,x = name)) + 
  geom_violin() + 
  geom_boxplot(width = .2) +
  labs(x="", y = "Point Winning Percentage")+ 
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_x_discrete(labels = c('First Serve','Second Serve w/o Double Faults','Second Serve w/ Double Faults'))
```


---

## <center> Players with a substantial Minutes Played advantage are favored in the finals </center>

```{r, echo=FALSE}
final_minutes%>%
  mutate(winner_adv = as.factor(winner_adv),
         winner_adv = ifelse(winner_adv == "1", 
                             "Minute Advantage", 
                             "Minute Disadvantage"),
         adv_category = fct_relevel(adv_category,
                                    "0-49",
                                    "50-99",
                                    "100-149",
                                    "150+"))%>%
  ggplot(aes(x = adv_category, fill = winner_adv))+
  geom_bar(position = "dodge")+
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))+
  # ggthemes::scale_fill_colorblind()+
  scale_fill_manual(values = c("#157394", "#dc4c4c"))+
  theme_bw()+
  labs(fill = "Winner Type",
       x = "Minute Differential",
       y = "Number of Final Matches") +
  theme(legend.position = "right",
        legend.title = element_text(face = "bold"),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        axis.ticks.x = element_blank())
```

---



## <center> Minimax clustering was performed to analyze serve aggression </center>

- Variables of interest were standardized: **aces** and **double faults**

- Minimum of 50 matches required

- k = 4 clusters chosen using:
  - **Dendrogram:** less than 4 clusters $\rightarrow$ large gap in dissimilarity between clusters 

--
  - **Context:** it was hypothesized that there are 4 categories of service players: 

      1. High aces, low double faults (best combination)
      2. High aces, high double faults (aggressive)
      3. Low aces, low double faults (conservative)
      4. Low aces, high double faults (worst combination)
      
- **Minimax** was chosen to identify cluster prototypes


---

## <center>Minimax clustering reveals a small cluster of elite players</center>

```{r, echo = FALSE, fig.height = 4}

player_serve_stats_filtered <- player_serve_stats %>%
  filter(matches_played >= 50) %>%
  mutate(std_aces_match = as.numeric(scale(aces_match)),
         # Standardizing aces and dfs
         std_df_match = as.numeric(scale(df_match))) 

wta_player_dist <- dist(dplyr::select(player_serve_stats_filtered, std_aces_match, std_df_match))

wta_minimax <-
  protoclust(wta_player_dist)

# Add in clusters to dataset

minimax_wta_player_clusters <- protocut(wta_minimax, k = 4)

clusters_stats_minimax <- player_serve_stats_filtered %>%
  mutate(player_clusters = 
           as.factor(minimax_wta_player_clusters$cl))


# Protos only dataset
wta_protos <- clusters_stats_minimax%>%
  slice(minimax_wta_player_clusters$protos)

# Clusters plotted with protos
clusters_stats_minimax%>%
  select(player_name, contains("match"), player_clusters)%>%
  ggplot(aes(x = aces_match, y = df_match, color = player_clusters))+
  geom_point(alpha = 0.5, size = 3)+
  geom_label(data = wta_protos, aes(label = player_name), size = 3)+
  ggthemes::scale_color_colorblind()+
  theme_bw()+
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))+
  labs(x = "Aces per Match",
       y = "Double Faults per Match",
       color = "Player Cluster",
       # title = "Minimax Clustering Players Based on Aces and Double Faults",
       # subtitle = "There are few elite players who have high aces and low double faults",
       caption = "Labeled players are\ncluster prototypes")

```

---

## Conclusions

1. Winning players faced fewer break points AND saved them at greater rates

2. Faulting on a first serve leads to a considerable drop in point winning percentage

3. Tournament finalists who have a 100+ minute advantage entering the final fare much better in the final match 

--

## Limitations

1. Only 221 tournaments included in the dataset, and only ~60 have a minute differential greater than a 100 minutes  

2. Lack of advanced tracking data (e.g. serve velocity, distance covered in a match)


---

## <center>Complete linkage leads to a wide cluster of elite players</center>

```{r, echo = FALSE, fig.height = 4}

wta_complete_hclust <- 
  hclust(wta_player_dist, method = "complete")

clusters_stats <- player_serve_stats_filtered %>%
  mutate(player_clusters = 
           as.factor(cutree(wta_complete_hclust, k = 4)))


# Key players to label in graph
key_players <- clusters_stats %>%
  filter(player_name %in% c("Serena Williams", 
                            "Maria Sharapova", 
                            "Naomi Osaka",
                            "Venus Williams"))

clusters_stats %>%
  select(player_name, contains("match"), player_clusters) %>%
  ggplot(aes(x = aces_match, y = df_match, color = player_clusters))+
  geom_point(alpha = 0.5, size = 3)+
  geom_label(data = key_players, aes(label = player_name), size = 3)+
  ggthemes::scale_color_colorblind()+
  theme_bw()+
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))+
  labs(x = "Aces per Match",
       y = "Double Faults per Match",
       color = "Player Cluster",
       # title = "Complete Linkage Clustering Players Based on Aces and Double Faults",
       # caption = "Minimum 50 matches played"
       #subtitle = "There are few elite players who have high aces and low double faults",
       )


```
